Last glacial cycle glacier erosion potential in the Alps

### 2021-02-04 articles/erosion/alps-erosion.tex
Rework climate discussion with Ian's comments.

    The glacial landscape of the Alps has fascinated generations of explorers,
    artists, mountaineers and scientists with its diversity, including
    erosional features of all scales from high-mountain cirques, to steep
    glacial valleys and large over-deepened basins. Using previous glacier
    modelling results, and empirical inferences of bedrock erosion under modern
    glaciers, we compute a distribution of potential glacier erosion in the Alps
    over the last glacial cycle from 120\,000 years ago to the present.
    %
    Despite large uncertainties pertaining to the climate history of the Alps and
    unconstrained glacier erosion processes, the resulting modelled patterns of
    glacier erosion includes persistent features. The cumulative imprint of
    the last glacial cycle shows a very strong localization of glacier erosion
    with local maxima at the mouths of major Alpine valleys and some other
    upstream sections where glaciers are modelled to have flown with the
    highest velocity. The modelled erosion rates vary significantly through the
    glacial cycle, but show paradoxically little relation to the total glacier
    volume. Phases of glacier advance and maximum extension, characterized by
    low air and glacier temperatures, see a localization of rapid erosion rates
    at low elevation, while the intra-montane glacial landscape is modelled to
    date from phases of intermediate glacier cover. The modelled erosion rates
    peak during deglaciation phases, when higher air and glacier temperature
    enhance glacier dynamics, implying that climatic conditions that result in
    rapid glacier erosion might be quite specific.
    %
    Our results depict the Alpine glacier erosion landscape as a
    time-transgressive patchwork, with different erosional landscapes of the
    same mountain range corresponding to different glaciation stages and time
    periods, thus offering insight into its diversity.

@@

    The ice-sheet model set-up was presented in an earlier publication
    \citep{Seguinot.etal.2018} and is briefly summarized here. The simulations
    were conducted using the Parallel Ice Sheet Model (PISM, development
    version~e9d2d1f), an open-source, finite-difference, shallow-ice and
    shallow-shelf model
    \citep{PISM-authors.2017}. The model includes temperature and water-content
    dependent creep, pseudo-plastic basal sliding accounting for till
    dilatation under high water pressure, bedrock deformation under the ice
    load, and a~positive degree-day (PDD) surface mass balance model. The model
    is initialized with assumed present-day ice thickness and equilibrium
    ice and bedrock temperature at 120\,ka, and ran to the present.
    The ice-sheet model physical parameters are listed in the earlier
    open-access publication \citep{Seguinot.etal.2018}, and the complete PISM
    set-up is stored in long-term archived model output metadata for the
    present study \citep{Seguinot.2021}, and for the earlier publication
    \citep{Seguinot.2020, Seguinot.2020a}.

@@

% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
 \subsection{Deglaciation rapid erosion}
 \label{sect:deglac_ero}
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
 \ian{This section title strange now, maybe \textbf{Rapid erosion during
     deglaciation}. However, what about this?  \textbf{Implications for
     understanding the relationship between erosion and climate}... I think
   it is not only the deglaciation that is interesting, but also the periods before.}

    The modelled, domain-integrated glacier erosion rates do not increase
    together with ice volume. On the opposite, periods of ice advance and
    maximum glaciation correspond to comparatively slow erosion
    (Figs.~\ref{fig:cumulative} and~\ref{fig:evolution}). While this result may
    seem paradoxical, and is modulated by the exponent, $l$, in the erosion
    power-law, it is in fact compatible with previous field-based studies that
    have evidenced an \ian{\textbf{do you mean inverse or direct? inverse}}
    relationship between air temperatures and glacier erosion rates
    \ian{in some cases} \citep{Koppes.etal.2015, Cook.etal.2020}.

    \ian{Cold temperatures generally reduce sliding \citep{Koppes.etal.2015} and
      if the glacier becomes cold-based then sliding, and thus erosion, will cease altogether.
      Furthermore, in cases of advance, ice surfaces gradients are lower and cold ice is
      advected downglacier, thus limiting erosion.}

    Deglaciation periods, characterised by a warmer climate and above-balance
    glacier extent, yield the highest domain-integrated erosion rates of the
    glacial cycle (Figs.~\ref{fig:cumulative} and~\ref{fig:evolution}).
    \ian{\textbf{I think it would be good to have a sentence about the
        differences in glacier dynamics that cause the increase.}
      Recent work, examining a denudation records, highlights that a non-linear forcing between
      climate and glacier erosion persists \citep{Mariotti.etal.2021}. This paper
      noted an increase in glacier erosion in the southern Alps, in the Var Valley,
      roughly at $25$\,\unit{Ka} Bp following an extended period of low dunudation rates.
      The increase in erosion here, predates the increase observed in our study
      ($\sim 15$\,\unit{Ka}; Figs.~\ref{fig:evolution}). However, similar processes caused by climate warming
      likely persist. This sedimentary record \citep{Mariotti.etal.2021} and the model results
      here demonstrate that climatic conditions over-which glacier erosion occurs
      are relatively specific and likely transient. }

    \ian{I added some citations here. Increases in erosion following glacier retreat
      seems to happen. But there are good critisims of the sedimentary record, and the mechanism
      seems a little bit uncertain. This is one of the cool things about this paper, but also makes this paragarph
    a bit delicate.}
    This
    result is compatible with the field-based studies, which have evidenced
    increased glacier erosion and pro-glacial sedimentation during deglaciation
    periods \citep[e.g.][]{Koppes.Montgomery.2009,Micheletti.Lane.2016,Lane.etal.2017}, including the ongoing glacier retreat
    \citep[e.g.][]{Bendixen.etal.2017}. Yet this increase has often been attributed
    to higher meltwater availability, promoting sediment transport \citep{Delaney.Adhikari.2020},
    enhanced glacier sliding \citep[][This citation might a bit strange]{Herman.etal.2011}, and hydrofracturing
    (plucking) \citep[][This citation is more about monthly to diurnal variability]{Ugelvig.etal.2018}.

    But importantly, these hydrological processes are not included in our
    model, so that deglaciation surface meltwater is assumed to instantly exit
    the glacier with no effect on glacier dynamics \ian{ \citep[e.g][]{Werder.etal.2013}
      or glacier erosion  \citep[][]{Iverson.2012,Ugelvig.etal.2018}}.
    The enhanced glacier
    velocity during periods of ice retreat is here modelled to result from
    changes of air temperature alone, propagating through diffusion and
    advection into the ice, increasing the glacier's internal energy through
    temperature and englacial water content, thereby facilitating shear strain
    and eventually activating basal sliding \citep[cf.][]{Aschwanden.etal.2012,
    Seguinot.etal.2016}. Observed feedbacks between surface meltwater and basal
    sliding, not included in the model, presumably further accelerated erosion
    during deglaciation \citep{Herman.etal.2011}. \ian{Furthermore, we omit the
      fluvial transport of sediment, which could well increase sediment discharge
      and erosion with increased melt, so long as additional sediment remains
      available \citep{Delaney.Adhikari.2020}. Rapid glacier erosion can be expected
    to accompany the  ongoing 21st-century deglaciation phase. This process might be
    especially strong on the ice sheets of Greenland and Antarctica where
    polythermal and cold-based glaciers are warming and sliding is increasing
    \citep[e.g.][]{Moon.etal.2012,Overeem.etal.2017}.}

>>>>

    The glacial landscape of the Alps has fascinated generations of explorers,
    artists, mountaineers and scientists with its diversity, including
    erosional features of all scales from high-mountain cirques, to steep
    glacial valleys and large over-deepened basins. Using previous glacier
    modelling results, and empirical inferences of bedrock erosion under modern
    glaciers, we compute a distribution of potential glacier erosion in the Alps
    over the last glacial cycle from 120\,000 years ago to the present.
    %
    Despite large uncertainties pertaining to the climate history of the Alps and
    unconstrained glacier erosion processes, the resulting modelled patterns of
    glacier erosion includes persistent features. The cumulative imprint of
    the last glacial cycle shows a very strong localization of glacier erosion
    with local maxima at the mouths of major Alpine valleys and some other
    upstream sections where glaciers are modelled to have flown with the
    highest velocity. The modelled erosion rates vary significantly through the
    glacial cycle, but show paradoxically little relation to the total glacier
    volume. Phases of glacier advance and maximum extension, characterized by
    low air and glacier temperatures, see a localization of rapid erosion rates
    at low elevation, while the intra-montane glacial landscape is modelled to
    date from phases of intermediate glacier cover. The modelled erosion rates
    peak during deglaciation phases, when higher air and glacier temperature
    enhance glacier dynamics, implying that climatic conditions that result in
    rapid glacier erosion might be quite transient and specific.
    %
    Our results depict the Alpine glacier erosion landscape as a
    time-transgressive patchwork, with different erosional landscapes of the
    same mountain range corresponding to different glaciation stages and time
    periods, thus offering insight into its diversity.

@@

    The ice-sheet model set-up was presented in an earlier publication
    \citep{Seguinot.etal.2018} and is briefly summarized here. The simulations
    were conducted using the Parallel Ice Sheet Model (PISM, development
    version~e9d2d1f), an open-source, finite-difference, shallow-ice and
    shallow-shelf model
    \citep{PISM-authors.2017}. The model includes temperature and water-content
    dependent creep, pseudo-plastic basal sliding accounting for till
    dilatation under high water pressure, bedrock deformation under the ice
    load, and a~positive degree-day (PDD) surface mass balance model. The model
    is initialized with assumed present-day ice thickness and equilibrium
    ice and bedrock temperature at 120\,ka (thousand years before the present),
    and ran to the present.
    The ice-sheet model physical parameters are listed in the earlier
    open-access publication \citep{Seguinot.etal.2018}, and the complete PISM
    set-up is stored in long-term archived model output metadata for the
    present study \citep{Seguinot.2021}, and for the earlier publication
    \citep{Seguinot.2020, Seguinot.2020a}.

@@

% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
\subsection{Climate control on erosion}
% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -

    \note{What about this new section title?}

    The modelled, domain-integrated glacier erosion volume do not increase
    together with ice volume. On the opposite, periods of ice advance and
    maximum glaciation correspond to comparatively slow erosion
    (Figs.~\ref{fig:cumulative} and~\ref{fig:evolution}). While this result may
    seem paradoxical, and is modulated by the exponent, $l$, in the erosion
    power-law, it is in fact compatible with previous field-based studies that
    have evidenced an increase of erosion rates with air temperatures for at
    least some sets of glaciers \citep{Koppes.etal.2015, Cook.etal.2020}.

    Cold temperatures generally reduce sliding \citep{Koppes.etal.2015} and if
    the glacier becomes cold-based then sliding, and thus erosion, will cease
    altogether. In cases of glacier advance, newly formed, atmosphere-cold ice
    expands onto cooled bedrock hindering ice deformation and basal sliding.
    \note{Ian: ``Furthermore, in cases of advance, ice surfaces
          gradients are lower and cold ice is advected downglacier, thus
          limiting erosion''. Is this also valid for relatively low-sloping
          piedmont glaciers?}
    On the other hand, deglaciation periods are characterised by a warmer
    climate and above-balance glacier extent. Glacier ice formed an isolating
    blanket from the atmosphere so that geothermal heat has had time to warm up
    the lower layers and activate the double-feedback loop between ice
    temperature, deformation heat dissipation and basal melt-induced sliding.
    Warmer air temperatures contribute to enhance deformation and sliding even
    more, yielding the highest domain-integrated erosion rates of the
    glacial cycle (Figs.~\ref{fig:cumulative} and~\ref{fig:evolution}).

    \note{I took the freedom to shorten the Var story quite a bit.}
    This result is corroborated by a recent denudation record from the
    Mediterranean Alps, which includes an increase in glacier erosion in the
    Var Valley, roughly at 25\,ka following an extended period of low
    dunudation rates \citep{Mariotti.etal.2021}. While that increase in
    erosion predates the increase observed in our study ($\sim$ 15\,ka;
    Fig.~\ref{fig:evolution}), it demonstrates that climatic conditions
    over-which glacier erosion occurs are relatively specific and likely
    transient. Other field-based studies have evidenced increased glacier
    erosion and pro-glacial sedimentation during deglaciation periods
    \citep[e.g.,][]{Koppes.Montgomery.2009, Micheletti.Lane.2016,
    Lane.etal.2017}, including the ongoing glacier retreat
    \citep[e.g.,][]{Bendixen.etal.2017}. Yet this increase has often been
    attributed to higher meltwater availability, promoting sediment transport
    \citep{Delaney.Adhikari.2020}, enhanced glacier sliding
    \citep{Herman.etal.2011}, and hydrofracturing (plucking)
    \citep{Hallet.1996, Ugelvig.etal.2018, Hildes.etal.2004}.

    But importantly, these hydrological processes are not included in our
    model, so that deglaciation surface meltwater is assumed to instantly exit
    the glacier with no effect on glacier dynamics \citep[cf.][for comparison]
    {Werder.etal.2013, Iverson.2012, Ugelvig.etal.2018}. The enhanced glacier
    velocity during periods of ice retreat is here modelled to result from
    changes of air temperature alone, propagating through diffusion and
    advection into the ice, increasing the glacier's internal energy through
    temperature and englacial water content, thereby facilitating shear strain
    and eventually activating basal sliding \citep[cf.][for the physics]
    {Aschwanden.etal.2012, Seguinot.etal.2016}. Observed feedbacks between
    surface meltwater and basal sliding, not included in the model, presumably
    further accelerated erosion during deglaciation \citep{Herman.etal.2011}.
    Furthermore, we omit the fluvial transport of sediment, which could well
    increase sediment discharge and erosion with increased melt, so long as
    additional sediment remains available \citep{Delaney.Adhikari.2020}. Rapid
    glacier erosion can be expected to accompany the ongoing 21st-century
    deglaciation phase, especially for the ice sheets of Greenland and
    Antarctica where polythermal and cold-based glaciers are warming and
    sliding is increasing \citep[e.g.,][]{Moon.etal.2012,Overeem.etal.2017}.
